I Compensate, therefore I Am \\ (accurate for DNN verification)

### 2024-01-19 Paper/Paper_noProof.tex
background

Given an input vector $\boldsymbol{z}^0 \in \mathbb{R}^{d_0}$, 
denoting $\hat{\boldsymbol{z}}^{0}={\boldsymbol{z}}^0$, we define inductively the value vector $\boldsymbol{z}^i$ at layer $1 \leq i \leq \ell$ with
\begin{align*}
	{\boldsymbol{z}}^{i} &= \boldsymbol{W}^i\cdot \hat{\boldsymbol{z}}^{(i-1)}+ \vb^i\\
	\hat{\boldsymbol{z}}^{i} &= ReLU({\boldsymbol{z}}^i).
\end{align*}

The vector $\hat{\boldsymbol{z}}$ is called post-activation values, 
$\boldsymbol{z}$ is called pre-activation values, 
and $\boldsymbol{z}^{i}_j$ is used to call the $j$-th neuron in the $i$-th layer. 
For $\boldsymbol{x}=\vz^0$ the vector of input, we denote by $f(\boldsymbol{x})=\vz^\ell$ the output. Finally, neurons are also called \emph{nodes}, and when we refer to a specific neuron, we use $a,b,c,d,n$ to denote them, and $W_{a,b} \in \mathbb{R}$ to denote the weight from neuron $a$ to $b$. Similarly, for input $\boldsymbol{x}$, we denote by $\val_{\boldsymbol{x}}(a)$ the value of neuron $a$ when the input is $\boldsymbol{x}$. A path $\pi$ is a sequence $\pi=(a_i)_{k \leq  i \leq k'}$ of neurons in consecutive layers, and the weight of $\pi$ is 
$weight(\pi)=W_{a_k,a_{k+1}} \times \cdots \times  W_{a_{k'-1},a_{k'}}$.

@@

Concerning the verification problem, we focus for simplicity on the usual local robustness question. Local robustness asks to determine whether the output of a neural network will be affected under small perturbations to the input. 
Formally, for an input $\vx$ perturbed by $\varepsilon >0$ under distance $d$, then the DNN is locally $\varepsilon$-robust in $\vx$ whenever:
\begin{align*}
	\forall \boldsymbol{x'} \text{ s.t. } d(x,x')\leq \varepsilon, \text{ we have }  
	argmax_i (f(\boldsymbol{x'})[i]) = argmax_i(f(\boldsymbol{x})[i])
\end{align*}

@@

In this paper, we consider $L^{\infty}$ norm the max value of distance of each dimension, that is $d(\boldsymbol{x},\boldsymbol{x}_0)=\max |\boldsymbol{x}(n)-\boldsymbol{x}_0(n)|$. 
\fi

@@

In this section, we describe different value (over-)abstractions that are used by efficient algorithms to certify the networks. Over-abstractions ensure that all values in the original network are possible in the over-abstraction, and thus if we prove safety in the over-abstraction, then we will have a proof of safety for the original image.

\subsection{Box and DeepPoly Abstraction}

@@

			\node[circle, draw= black, thick, minimum width = 20,
			minimum height = 20] (input1) {$x_1$};

			\node[circle, draw= purple, thick, minimum width = 20,
			minimum height = 20] (input2) at ($(input1) + (0,-1.5)$) {$x_2$};

@@

			\node[circle, draw= black, thick, minimum width = 20,
			minimum height = 20] (hidden1) at ($(input1) + (2.5,0)$) {$x_3$};
			\node[circle, draw= black, thick] (hidden2) at ($(input1) + (2.5,-1.5)$) {$x_4$};

			\node[circle, draw= black, thick, minimum width = 20,
			minimum height = 20] (hidden3) at ($(input1) + (5,0)$){$\hat{x}_3$};
			\node[circle, draw= black, thick] (hidden4) at ($(input1) + (5,-1.5)$) {$\hat{x}_4$};

			\node[circle, draw= blue, thick, minimum width = 20,
			minimum height = 20] (hidden5) at ($(input1) + (7.5,0)$){$x_5$};
			\node[circle, draw= black, thick] (hidden6) at ($(input1) + (7.5,-1.5)$) {$x_6$};

			% Output layer
			\node[circle, draw= black, thick, minimum width = 20,
			minimum height = 20] (output1) at ($(input1) + (10,0)$){$\hat{x}_5$};

			\node[circle, draw= black, thick, minimum width = 20,
			minimum height = 20] (output2) at ($(input1) + (10,-1.5)$){$\hat{x}_{6}$};

@@

			\draw[->,color=green, thick] (hidden1) -- (hidden3) node[midway, above] {$\max(x_1,0)$};
			\draw[->,color=red, thick] (hidden2) -- (hidden4) node[midway, above] {$\max(x_2,0)$};

@@

			\draw[->,thick] (hidden5) -- (output1) node[midway, above] {$\max(x_5,0)$};
			\draw[->,thick] (hidden6) -- (output2) node[midway, above] {$\max(x_6,0)$};

		\end{tikzpicture}
		\caption{A DNN example from \cite{kpoly}, where every node is separated into 2, $x$ after the weighted sum and $\hat{x}$ after the ReLU activation function. The pair $(x_2 x_3 x_5,x_2 x_4 x_5)$ in green and red is compensating (weights of paths are $1,-2$).}
		\label{fig1}
	\end{figure}

An efficient and prolific line of work concerning DNN verification has been to develop value abstraction of the network. The idea is to compute upper and lower bounds on the values of (some) neurons in the DNN when the inputs are in a given neighbourhood, in order to conclude on the robustness without needing to compute the values exactly for all inputs.

First, remark that weighted sums is a linear function, so it is easy to represent it explicitly. The only problematic part to represent accurately is the ReLU function. While it is fairly simple piecewise linear function with 2 modes ($x<0$ and $x \geq 0$), it is not linear. Considering its iterated impact along the different layers of a ReLU DNN is much more complex that it seems. First, let us remark that it is easy to represent exactly $\ReLU(x)$
in case $x$ is known to be entirely positive and in the case where it is entirely negative
(such a neuron $x$ is called stable), as in both case there is only one known linear mode involved. Hence, the sole problem remains how to treat unstable neurons.

Consider the simpler abstraction, called the {\em box abstraction} \cite{deeppoly}: from the bounds in the input, one compute by the linear step the bound on the next layer, apply the ReLU function, and so on. For instance, on the DNN example of Fig \ref{fig1}, $x_1,x_2 \in [-1,1]$, then $x_3,x_4 \in [-2,2]$, using ReLU $\hat{x}_3,\hat{x}_4 \in [0,2]$, and $x_5 \in [0,6]$ while $x_6 \in [0,2]$. 
While the bounds of $x_1, \ldots, x_4$ are exact (for all $\alpha$ 
in the range, there is an input $\vy$ such that $\val_{\vy}(x_i)=\alpha$), 
it is no more the case from the next layer (starting in $x_5,x_6$) because of potential 
dependencies between earlier neurons: $x_3$ reaches weight $2$ only when $x_1=x_2=2$,
in which case $x_4$ reaches weight $0$. Hence, it is impossible that $\Val(x_5)$ reaches $6$, as it would require both $x_3$ and $x_4$ to have weight $2$.

A still extremely fast algorithm that corrects some of this issue is known as {\em DeepPoly} \cite{deeppoly}, also rediscovered as the {\em CROWN} algorithm \cite{crown}. Instead of 
hard bounds, for each neuron $n$ of layer $k$, two affine functions (with input at the previous layer $k-1$) are kept as lower bound and upper bound of the weight of the node.
For instance, $f_i \leq \Val(\hat{x}_i) \leq g_i$ with 
$f_{3}(x_3)=f_4(x_4)=0$ and 
$g_3(x_3) = \frac{x_3+2}{2}$,
$g_4(x_4) = \frac{x_4+2}{2}$. 
With that, we get 
$x_5 \leq g_3(x_3) + 2 g_4(x_4) = \frac{x_3 + 2x_4 + 6}{2} = \frac{3x_1 - x_2 + 6}{2}\leq 5$.

While there is one optimal linear function for the upper bound $g_i(x_i)= \beta \frac{x_i-\alpha}{\beta-\alpha}$ for ReLU nodes, there are two choices for the lower bound: $f_i(x_i) = 0$ or $f_i(x_i)=x_i$.
In the DeepPoly algorithm, the choice is made by considering the range of an unstable neuron 
$x_i=[\alpha,\beta]$ with $\alpha < 0 < \beta$. 
If $|\alpha|\geq |\beta|$ then $0$ is used, while if $|\beta|>|\alpha|$ then $x_i$ (the identity function) is used. Let us call $\overline{\mbox{DeepPoly}}$ the DeepPoly algorithm such that 
$f_i(x_i) = 0$ is always used, and $f_i(x_i) = x_i$ is never used. 
It has a main advantage, as it subsumes the box abstraction.
Indeed, taking an interval $x_i \in [-0.2,5]$, it gives $\ReLU(x_i) \in [0,5]$, 
while the normal DeepPoly algorithm would only allow to conclude that $\ReLU(x_i) \in [-0.2,5]$.

@@

At the other end of the spectrum, we find the Mixed Integer Linear Programming (MILP) value abstraction \cite{MILP}, which is a priori a complete method (but whose complete encoding is much slower). 
Consider an unstable neuron $y \in[\alpha,\beta]$. The value of $\ReLU(y)$ can be encoded exactly in an MILP formula with one integer (actually even binary) variable $a$ valued in ${0,1}$, using 4 constraints:
\vspace{-0.1cm}
\begin{align*}
	\hat{y} \geq y \quad \wedge \quad \hat{y} \geq 0, \quad \wedge \quad \hat{y} \leq \beta \cdot a \quad \wedge \quad \hat{y} \leq y-\alpha \cdot (1-a)
\end{align*}

This encoding is exact in the meaning that \cite{MILP} for all $y \in [\alpha,\beta] \setminus 0$, there is a unique solution $(a,\hat{y})$ satisfying the constraints, and we have $\hat{y}=\ReLU(y)$ (and $a$ is 0 if $y < 0$ and 1 if $y>0$, and can be either for $y=0$). Such an encoding can be used for all (unstable) ReLU nodes, and optimizing its value allows to certify a given image. However, for any networks hundreds of nodes or more, it means having hundreds of integer variables, and such network will usually not be solvable efficiently.

@@

As far as we know, no formal result states the relationship between the LP approximation and DeepPoly.
We prove now that the LP abstraction is strictly more accurate that the DeepPoly abstraction, as it actually corresponds to considering in DeepPoly both linear functions lower bounding the ReLU function, $ReLU(y) \geq 0$ and $ReLU(y) \geq y$, while DeepPoly considers exactly one.

\begin{proposition}
	\label{LP}
 Given $y \in [\alpha,\beta]$ with $\alpha < 0 < \beta$, the following two systems of constraints 
 1) for LP and 2) for DeepPoly with both lower bounds are equivalent:
 \vspace{-0.3cm}
 \begin{align*}
	 & 1) \quad \hat{y} \geq y \quad \wedge \quad \hat{y} \geq 0, \quad \wedge \quad \hat{y} \leq \beta \cdot a \quad \wedge \quad \hat{y} \leq y-\alpha \cdot (1-a), \, a \in [0,1] \\
	 \text{and} \quad  & 2)  \quad \hat{y} \geq y \quad \wedge \quad \hat{y} \geq 0 \quad \wedge \quad \hat{y} \leq \beta \frac{y-\alpha}{\beta-\alpha}
\end{align*} 
\end{proposition}

@@

It remains to prove that the 2 upper bound constraints of 1) LP corresponds to the 2) DeepPoly linear upper bound. For that, we study the upper bound function in the variable $a \in  [0,1]$, with $y \in [\alpha,\beta]$ fixed. We have $\hat{y}$ in 1) upper bounded by $max_{a \in [0,1]} (min(\beta \cdot a, y - \alpha (1-a)))$, and this bound can be reached. 
The function $min(\beta \cdot a, y - \alpha (1-a))$ reaches its maximum for 
$\beta \cdot a = y - \alpha (1-a)$, that is 
$(\beta - \alpha) a = y - \alpha$, i.e. 
$$ a = \frac{y - \alpha}{\beta-\alpha}$$

It thus gives an upper bound at 
$\beta \cdot a = \beta \frac{y - \alpha}{\beta-\alpha}$, which is exactly the function used in DeepPoly.
\hfill $\square$
\end{proof}

>>>>

Given an input vector $\boldsymbol{z}^0 \in \mathbb{R}^{d_0}$, 
denoting $\hat{\boldsymbol{z}}^{0}={\boldsymbol{z}}^0$, we define inductively the value vectors $\boldsymbol{z}^i,\hat{\vz}^i$ at layer $1 \leq i \leq \ell$ with
\begin{align*}
	{\boldsymbol{z}}^{i} &= \boldsymbol{W}^i\cdot \hat{\boldsymbol{z}}^{i-1}+ \vb^i\\
	\hat{\boldsymbol{z}}^{i} &= ReLU({\boldsymbol{z}}^i).
\end{align*}

The vector $\hat{\boldsymbol{z}}$ is called post-activation values, 
$\boldsymbol{z}$ is called pre-activation values, 
and $\boldsymbol{z}^{i}_j$ is used to call the $j$-th neuron in the $i$-th layer. 
For $\boldsymbol{x}=\vz^0$ the (vector of) input, we denote by $f(\boldsymbol{x})=\vz^\ell$ the output. Finally, pre- and post-activation neurons are called \emph{nodes}, and when we refer to a specific node/neuron, we use $a,b,c,d,n$ to denote them, and $W_{a,b} \in \mathbb{R}$ to denote the weight from neuron $a$ to $b$. Similarly, for input $\boldsymbol{x}$, we denote by $\val_{\boldsymbol{x}}(a)$ the value of neuron $a$ when the input is $\boldsymbol{x}$. A path $\pi$ is a sequence $\pi=(a_i)_{k \leq  i \leq k'}$ of neurons in consecutive layers, and the weight of $\pi$ is 
$weight(\pi)=W_{a_k,a_{k+1}} \times \cdots \times  W_{a_{k'-1},a_{k'}}$.

@@

Concerning the verification problem, we focus for simplicity on the usual local robustness question. Local robustness asks to determine whether the output of a neural network will be affected under small perturbations to the input. 
Formally, for an input $\vx$ perturbed by $\varepsilon >0$ under distance $d$, then the DNN is locally $\varepsilon$-robust in $\vx$ whenever:
\begin{align*}
	\forall \boldsymbol{x'} \text{ s.t. } d(\vx,\vx')\leq \varepsilon, \text{ we have }  
	argmax_i (f(\boldsymbol{x'})[i]) = argmax_i(f(\boldsymbol{x})[i])
\end{align*}

@@

In this paper, we consider $L^{\infty}$ norm the max value of distance of each dimension, that is $d(\vx,\boldsymbol{x}_0)=\max |\boldsymbol{x}(n)-\boldsymbol{x}_0(n)|$. 
\fi

@@

In this section, we describe different value (over-)abstractions on $\vz$ that are used by efficient algorithms to certify robustness around an input $\vx$. Over-abstractions of values include all values for $\vz$ in the neighbourhood of $\vx$, and thus a certificate for safety in the over-abstraction is a proof of safety for the original input $\vx$.

\subsection{The Box and DeepPoly Abstractions}

@@

			\node[circle, draw= black, thick, minimum width = 20,
			minimum height = 20] (input1) {$n_1$};

			\node[circle, draw= purple, thick, minimum width = 20,
			minimum height = 20] (input2) at ($(input1) + (0,-1.5)$) {$n_2$};

@@

			\node[circle, draw= black, thick, minimum width = 20,
			minimum height = 20] (hidden1) at ($(input1) + (2.5,0)$) {$n_3$};
			\node[circle, draw= black, thick] (hidden2) at ($(input1) + (2.5,-1.5)$) {$n_4$};

			\node[circle, draw= black, thick, minimum width = 20,
			minimum height = 20] (hidden3) at ($(input1) + (5,0)$){$\hat{n}_3$};
			\node[circle, draw= black, thick] (hidden4) at ($(input1) + (5,-1.5)$) {$\hat{n}_4$};

			\node[circle, draw= blue, thick, minimum width = 20,
			minimum height = 20] (hidden5) at ($(input1) + (7.5,0)$){$n_5$};
			\node[circle, draw= black, thick] (hidden6) at ($(input1) + (7.5,-1.5)$) {$n_6$};

			% Output layer
			\node[circle, draw= black, thick, minimum width = 20,
			minimum height = 20] (output1) at ($(input1) + (10,0)$){$\hat{n}_5$};

			\node[circle, draw= black, thick, minimum width = 20,
			minimum height = 20] (output2) at ($(input1) + (10,-1.5)$){$\hat{n}_{6}$};

@@

			\draw[->,color=green, thick] (hidden1) -- (hidden3) node[midway, above] {$\max(n_1,0)$};
			\draw[->,color=red, thick] (hidden2) -- (hidden4) node[midway, above] {$\max(n_2,0)$};

@@

			\draw[->,thick] (hidden5) -- (output1) node[midway, above] {$\max(n_5,0)$};
			\draw[->,thick] (hidden6) -- (output2) node[midway, above] {$\max(n_6,0)$};

		\end{tikzpicture}
		\caption{A DNN example from \cite{kpoly}: every neuron is separated into 2 nodes, $n$ pre- and $\hat{n}$ post-ReLU activation function. The pair $(n_2 n_3 n_5,n_2 n_4 n_5)$ in green and red is compensating (weights of paths are $1,-2$).}
		\label{fig1}
	\end{figure}

The idea to perform value abstraction is to compute upper and lower bounds on the values of (some) neurons in the DNN when the inputs are in a given neighbourhood, in order to conclude on the robustness without needing to compute the values exactly for all inputs (in the neighbourhood).

First, remark that weighted sums is a linear function, so it is easy to represent it explicitly. The only problematic part to represent accurately is the ReLU function. While it is fairly simple piecewise linear function with 2 modes ($x<0$ and $x \geq 0$), it is not linear. Considering its iterated impact along the different layers of a ReLU DNN is much more complex than it seems. First, let us remark that it is easy to represent exactly $\ReLU(x)$ in case $x$ is {\em stable}, that is, either it is known to be entirely positive, order
it is known to be entirely negative, as in both case there is only one known linear mode involved. Hence, the sole problem remains how to handle {\em unstable} neurons.

Consider the simpler abstraction, called the {\em Box abstraction} \cite{deeppoly}: inductively compute the bound on each neuron of the next layer independently, by considering the weighted sum over the bounds of the previous layer, then clip the lower bound at 
$\max(0,$ lower bound$)$ to represent the ReLU function, and so on.
For all $i$, let $x_i=\val_{\vx}(n_i)$, with $\vx=(x_1,x_2)$.
For instance, on the DNN example of Fig \ref{fig1}, $x_1,x_2 \in [-1,1]$, then $x_3,x_4 \in [-2,2]$, using ReLU $\hat{x}_3,\hat{x}_4 \in [0,2]$, and $x_5 \in [0,6]$ while $x_6 \in [0,2]$. 
While the bounds for $n_1, \ldots, n_4$ are exact (for all $\alpha$ 
in the range, there is an input $\vy$ such that $\val_{\vy}(n_i)=\alpha$), 
it is no more the case from the next layer (starting in $n_5,n_6$) because of potential 
dependencies between earlier neurons: $n_3$ reaches value $2$ only when $x_1=x_2=2$,
in which case $n_4$ reaches value $Val_{(2,2)}(n_4)=0$. Hence, it is impossible that $x_5=\Val_{\vx}(n_5)$ reaches $6$, as it would require both $n_3$ and $n_4$ to have value $2$.

A still extremely fast algorithm that corrects some of this issue is {\em DeepPoly} \cite{deeppoly}, also rediscovered as the {\em CROWN} algorithm \cite{crown}. Instead of 
hard bounds, for each neuron $n$ of layer $k$, two affine functions (with input at the previous layer $k-1$) are kept as lower bound and upper bound of the value of the node.
Denote $f_i \leq x_i \leq g_i$ with e.g.  
$f_{3}(x_3)=f_4(x_4)=0$ and 
$g_3(x_3) = \frac{x_3+2}{2}$,
$g_4(x_4) = \frac{x_4+2}{2}$. 
With that, we get 
$x_5 \leq g_3(x_3) + 2 g_4(x_4) = \frac{x_3 + 2x_4 + 6}{2} = \frac{3x_1 - x_2 + 6}{2}\leq 5$.

For $[\alpha,\beta]$ bounds on $x_i$, there is one optimal linear function for the upper bound $g_i(x_i)= \beta \frac{x_i-\alpha}{\beta-\alpha}$ for ReLU nodes.
There are two choices for the lower bound: $f^1_i(x_i) = 0$ or $f^2_i(x_i)=x_i$.
In the DeepPoly algorithm, the choice is made by considering $\alpha < 0 < \beta$ (for unstable neurons). If $|\alpha|\geq |\beta|$ then $f_i=f^1_i$ is used, while if $|\beta|>|\alpha|$ then $f_i=f^2_i$ is used. Denote {\em $\overline{\mbox{DeepPoly}}$} the DeepPoly algorithm such that $f^1_i$ is always used, and $f^2_i$ is never used. 
Unlike DeepPoly, {\em $\overline{\mbox{DeepPoly}}$} subsumes the {\em Box abstraction}. Indeed, taking bounds $[-0.2,5]$ on $x_i$, it gives $\ReLU(x_i) \in [0,5]$, 
while the normal DeepPoly algorithm would only allow to conclude that $\ReLU(x_i) \in [-0.2,5]$.

@@

At the other end of the spectrum, we find the Mixed Integer Linear Programming (MILP) value abstraction, which is a priori a complete method (albeit much slower than DeepPoly). 
Consider an unstable neuron $n \in[\alpha,\beta]$. The value $x$ of $\ReLU(n)$ can be encoded exactly in an MILP formula with one integer (actually even binary) variable $a$ valued in ${0,1}$, using 4 constraints \cite{MILP}:
\vspace{-0.1cm}
\begin{align*}
	\hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0, \quad \wedge \quad \hat{x} \leq \beta \cdot a \quad \wedge \quad \hat{x} \leq x-\alpha \cdot (1-a)
\end{align*}

From \cite{MILP}, for all $x \in [\alpha,\beta] \setminus 0$, there is a unique solution $(a,\hat{x})$ satisfying the constraints, and we have $\hat{x}=\ReLU(x)$ (and $a$ is 0 if $x < 0$ and 1 if $x>0$, and can be either for $x=0$). Such an encoding can be used for every (unstable) ReLU node, and optimizing its value allows to certify a given input. 
For Networks with hundreds of nodes or more, the associated MILP formula will have have hundreds of integer variables, and it will usually not be solvable efficiently.

@@

As far as we know, no formal result states the relationship between the LP approximation and DeepPoly.
We prove now that the LP abstraction is strictly more accurate that the DeepPoly abstraction, as it actually corresponds to considering in DeepPoly both linear functions lower bounding the ReLU function, $ReLU(n) \geq 0$ and $ReLU(n) \geq n$, while DeepPoly considers exactly one.

\begin{proposition}
	\label{LP}
 Given $n \in [\alpha,\beta]$ with $\alpha < 0 < \beta$, the following two systems of constraints 
 1) for LP and 2) for DeepPoly with both lower bounds are equivalent:
 \vspace{-0.3cm}
 \begin{align*}
	 & 1) \quad \hat{n} \geq n \quad \wedge \quad \hat{n} \geq 0, \quad \wedge \quad \hat{n} \leq \beta \cdot a \quad \wedge \quad \hat{n} \leq n-\alpha \cdot (1-a), \, a \in [0,1] \\
	 \text{and} \quad  & 2)  \quad \hat{n} \geq n \quad \wedge \quad \hat{n} \geq 0 \quad \wedge \quad \hat{n} \leq \beta \frac{n-\alpha}{\beta-\alpha}
\end{align*} 
\end{proposition}

@@

It remains to prove that the 2 upper bound constraints of 1) LP corresponds to the 2) DeepPoly linear upper bound. For that, we study the upper bound function in the variable $a \in  [0,1]$, with $n \in [\alpha,\beta]$ fixed. We have $\hat{n}$ in 1) upper bounded by $max_{a \in [0,1]} (min(\beta \cdot a, n - \alpha (1-a)))$, and this bound can be reached. 
The function $min(\beta \cdot a, n - \alpha (1-a))$ reaches its maximum for 
$\beta \cdot a = n - \alpha (1-a)$, that is 
$(\beta - \alpha) a = n - \alpha$, i.e. 
$$ a = \frac{n - \alpha}{\beta-\alpha}$$

It thus gives an upper bound at 
$\beta \cdot a = \beta \frac{n - \alpha}{\beta-\alpha}$, which is exactly the function used in DeepPoly.
\hfill $\square$
\end{proof}
